Learning complex models with invertible neural networks: a likelihood-free Bayesian approach

### 2019-06-24 latex/paper_main.tex
new latex files

\dropcap{M}athematical models are formal descriptions of scientific theories allowing a clear and unambiguous way to formulate and test scientific hypotheses about probabilistic phenomena in a probabilistic world. In its most abstract form, a mathematical model is specified by a set of parameters $\boldsymbol{\theta}$ and a generative model $q$ mimicking the process by which manifest quantities $\boldsymbol{x}$ arise from latent parameters: 
\begin{align*}
\boldsymbol{x}= q(\boldsymbol{\theta}) \numberthis \label{eqn:1} 
\end{align*}
While $q$ can represent an arbitrarily complex process by an arbitrarily complicated expression, its functional form is usually guided by a well-founded theoretical framework. For instance, $q$ can be a stochastic differential equation describing the dynamics of single neurons in the brain, or a step-by-step mechanism dictating the rate of gene expression in certain cells. Thus, it is only through theoretical embedding that a meaningful interpretation in terms of some mechanism can be attached to the parameters of a mathematical model. Examples of mathematical models can be found in various scientific domains, e.g., genetics \cite{zappia2017splatter, beaumont2002approximate}, cognitive science \cite{palestro2018likelihood, usher2001time}, neuroscience \cite{hwang2018conditional, lueckmann2017flexible}, population dynamics \cite{wood2010statistical}, epidemiology \cite{hethcote2000mathematics}, just to name a few.

Once a mathematical model has been formulated, the next step consists of fitting the model to experimental or observational data and recovering the parameters of interest. However, estimating the parameters of a mathematical model can quickly become one of the most tenacious challenges in applications to real-world problems. It is also one of the most important ones to be tackled, since without reliable parameter estimation methods, it is impossible to test the utility of a model, regardless of its sophistication or theoretical appeal. Idealized parameter estimation involves computing the inverse (backward) model $\boldsymbol{\theta} = q^{-1}(\boldsymbol{x})$ exactly. However, due to noise, inherent stochasticity or loss of information, the inverse usually does not exist, so researchers need to resort to the sophisticated frameworks of Bayesian or frequentist inference.

@@

Likelihood-free methods aim at bypassing the above problems by resorting to a simulation-based approach to parameter estimation and model selection \cite{palestro2018likelihood, turner2014generalized}. A subset of likelihood-free methods includes approximate Bayesian computation (ABC) methods, which aim at preserving the advantages of Bayesian data analysis even when the likelihood function is intractable or practically impossible to compute \cite{turner2014generalized, sunnaaker2013approximate, csillery2010approximate}. ABC methods approximate the likelihood function by repeatedly sampling parameters from a pre-specified prior distribution $\pi(\boldsymbol{\theta})$ and then simulating multiple datasets by running the generative model $q(\boldsymbol{\theta})$ using the sampled parameters. Thus, the core ingredients of ABC methods are a prior on $\boldsymbol{\theta}$, and a generative model $q(\boldsymbol{\theta})$, usually specified as a function code in a general-purpose programming language \cite{csillery2010approximate, mertens2018abrox}.

@@

Recently, ideas from machine learning and deep learning research have entered the field of likelihood-free inference in an attempt to overcome some of the shortcomings of traditional methods \cite{radev2019towards, hwang2018conditional, mestdagh2018prepaid, raynal2018abc, jiang2017learning, lueckmann2017flexible, papamakarios2016fast}. The most common approach has been to cast the problem of parameter estimation as a supervised learning task. In this setting, a large dataset of the form $\boldsymbol{D} = \{h(\boldsymbol{x}^{(i)}), \boldsymbol{\theta}^{(i)}\}_{i=1}^{n}$ is created by repeatedly sampling from $\pi(\boldsymbol{\theta})$ and simulating an artificial dataset $\boldsymbol{x}$ by running $q(\boldsymbol{\theta})$ with the sampled parameters. Usually, the dimensionality of the simulated data is reduced by computing summary statistics  with a fixed summary function $h(\boldsymbol{x})$. Then, a supervised learning algorithm $f(h(\boldsymbol{x});\boldsymbol{\phi}) = \widehat{\boldsymbol{\theta}}$ with learnable parameters $\boldsymbol{\phi}$ (e.g., linear regression, random forest, neural network) is trained on the simulated data to later output an estimate of the true data generating parameters. Thus, $f(h(\boldsymbol{x});\boldsymbol{\phi})$ essentially attempts to “learn” the intractable inverse model $\boldsymbol{\theta} = q^{-1}(\boldsymbol{x})$.

@@

In the context of mathematical modeling and Bayesian inference, the target distribution is the posterior distribution of model parameters $p(\boldsymbol{\theta}|\boldsymbol{x})$ capturing our uncertainty about the numerical values of parameters given empirical data. We can leverage the fact that most mathematical models are generative in nature and as such can be used to perform multiple simulations of the process of interest. By specifying a prior distribution over the model parameters $\pi(\boldsymbol{\theta})$, one can generate arbitrarily large datasets of the form $\boldsymbol{D} = \{\boldsymbol{x}^{(i)}, \boldsymbol{\theta}^{(i)}\}_{i=1}^{n}$ and use a deep generative model to learn a probabilistic mapping from data to parameters. Thus, at inference time, one can condition the model on observed data $\boldsymbol{x}_{obs}$ and obtain samples $\boldsymbol{\theta}_{1}, \boldsymbol{\theta}_{2},...,\boldsymbol{\theta}_{L}$  approximating the posterior $p(\boldsymbol{\theta}|\boldsymbol{x}=\boldsymbol{x}_{obs})$ in the manner described above.

@@

The basic building block of a cINN is the affine coupling block (ACB) \cite{ardizzone2018analyzing, kingma2018glow, dinh2016density}. Each ACB consists of four separate fully connected neural networks denoted as $s_{1}(\cdot), s_{2}(\cdot), t_{1}(\cdot), t_{2}(\cdot)$. An ACB is specifically designed to be invertible, which means that in addition to a parametric mapping $f_{\boldsymbol{\phi}_{inv}}: \mathbb{R}^{d} \rightarrow \mathbb{R}^{d}$ is also learns the inverse mapping $f_{\boldsymbol{\phi}_{inv}}^{-1}: \mathbb{R}^{d} \rightarrow \mathbb{R}^{d}$ "for free". Denoting the input vector of $f_{\boldsymbol{\phi}_{inv}}$ as $\boldsymbol{u}$ and the output vector as $\boldsymbol{v}$, it follows that $f(\boldsymbol{u}; \boldsymbol{\phi}_{inv}) = \boldsymbol{v}$ and $f^{-1}(\boldsymbol{v}; \boldsymbol{\phi}_{inv}) = \boldsymbol{u}$. Invertibility is achieved by splitting the input vector into two parts $\boldsymbol{u} = (\boldsymbol{u}_{1}, \boldsymbol{u}_{2})$ and performing the following operations on the split input:
\begin{align*} 
\boldsymbol{v}_{1} &= \boldsymbol{u}_{1} \odot \exp(s_{1}(\boldsymbol{u}_{2})) + t_{1}(\boldsymbol{u}_{2}) \numberthis \label{eqn:2}  \\ 
\boldsymbol{v}_{2} &= \boldsymbol{u}_{2} \odot \exp(s_{1}(\boldsymbol{v}_{1})) + t_{1}(\boldsymbol{v}_{1}) \numberthis \label{eqn:3} 
\end{align*}
The outputs $\boldsymbol{v} = (\boldsymbol{v}_{1}, \boldsymbol{v}_{2})$ are then concatenated again and passed to the next ACB. The inverse operation is given by:
\begin{align*} 
\boldsymbol{u}_{2} &= (\boldsymbol{v}_{2} - t_{2}(\boldsymbol{v}_{1})) \odot \exp(-s_{2}(\boldsymbol{v}_{1})) \numberthis \label{eqn:4}  \\ 
\boldsymbol{u}_{1} &= (\boldsymbol{v}_{1} - t_{1}(\boldsymbol{u}_{2})) \odot \exp(-s_{1}(\boldsymbol{u}_{2})) \numberthis \label{eqn:5} 
\end{align*}
An additional property of this design, which becomes relevant later for optimization, is that the operations of the ACB have tractable, and cheaply computable Jacobians (strictly upper or lower triangular matrices). Furthermore, the internal networks $s_{1}(\cdot), s_{2}(\cdot), t_{1}(\cdot), t_{2}(\cdot)$ can be represented by arbitrarily complex neural networks, which themselves need not be invertible, since they are only ever evaluated in the forward direction during both the forward and the inverse pass through the ACB. To ensure that the model is powerful enough to represent complicated distributions, we chain multiple ACBs, so that the output of each ACB becomes the input of the next. In this way, the whole chain remains invertible from the first input to the last output and can be viewed as a single function parameterized by trainable parameters $\boldsymbol{\phi}_{inv}$.

@@

\subsection*{Training the Networks}
We train all cINNs and summary networks described in this paper jointly via backpropagation. For all following experiments, we use the Adam optimizer  with a starter learning rate of $10^-3$ and an exponential decay rate of $.95$. We set the weight regularization parameter $\tau$ to a value of $10^{-5}$. Note, that we did not perform an extensive search for optimal values of the cINN parameter but stick to a cINN with 10 ACBs for each example in this paper.
Concerning the data generation step, we use two training approaches. The first follows the classical approximate Bayesian computation approach to create a large “reference table” or grid of the form $\boldsymbol{D} = \{\boldsymbol{x}^{(i)}, \boldsymbol{\theta}^{(i)}\}_{i=1}^{n}$. The reference table is then used as training data for the neural network and training continues for a pre-specified number of epochs through the entire reference table. A separate validation dataset is eventually used to assess the performance of the network. This training approach separates the simulation from the training phase but can incur large memory overhead, as the reference table must be stored on disk and then loaded in chunks or in its entirety into memory. The second approach follows a different strategy, which is used mainly in the field of active learning. Correspondingly, a dataset, or a batch of datasets, is created on the fly and then passed through the neural network. This training regime has the advantage that the network never “experiences” the same input data twice. Moreover, training can continue as long as the network keeps improving (i.e., the loss keeps decreasing), since overfitting in the classical sense is nearly impossible. However, if the simulations are computationally expensive and researchers need to experiment with multiple models, it might be beneficial to switch to the first regime, since simulation and training in the active learning regime are tightly intertwined. TODO - Describe for which!

@@

On an abstract level, our method requires three key ingredients: 1) a mathematical process model $q(\boldsymbol{\theta})$ capable of simulating data $\boldsymbol{x}$; 2) a prior distribution over the model parameters $\pi(\boldsymbol{\theta})$ encoding our prior beliefs about plausible parameter values; 3) an invertible neural network $f_{\boldsymbol{\theta}}$ capable of approximating a large enough family of probability distributions (see Figure 2). In practice, a chain of up to 10 ACBs should suffice to learn most distributions encountered in the life sciences, since they tend to be unimodal and relatively simple (in contrast to the distribution of natural images or words in a spoken language). From these three ingredients, a universal and reusable sampler can be designed for likelihood-free Bayesian estimation of both tractable and intractable mathematical models. \textbf{Algorithm} \ref{Algorithm:algo} describes the essential steps of the method using an arbitrary summary network and employing the active learning training regime.

\begin{algorithm}
\caption{Bayesian likelihood-free inference with invertible neural networks}\label{Algorithm:algo}
\begin{algorithmic}[1]
\State \emph{Training (via active learning):}
\Repeat
\State {Sample a batch of $\{\boldsymbol{\theta}^{(i)}\}_{i=1}^{m}$ from prior $\pi(\boldsymbol{\theta})$}
\State {Simulate a batch of datasets $\{\boldsymbol{x}^{(i)}\}_{i=1}^{m}$ by running $\boldsymbol{x}^{(i)} = q(\boldsymbol{\theta}^{(i)})$ for $i=1,...,m$}
\State {Pass $\{\boldsymbol{x}^{(i)}\}_{i=1}^{m}$ through summary network $h(\boldsymbol{x}^{(i)};\boldsymbol{\phi}_{sum})$ to obtain $\{\tilde{\boldsymbol{x}}^{(i)}\}_{i=1}^{m}$}
\State {Pass $\{\boldsymbol{\theta}^{(i)}\}_{i=1}^{m}$ and $\{\tilde{\boldsymbol{x}}^{(i)}\}_{i=1}^{m}$ through cINN $f(\boldsymbol{\theta}^{(i)};\boldsymbol{x}^{(i)},\boldsymbol{\phi}_{inv})$ to obtain $\{\boldsymbol{z}^{(i)}\}_{i=1}^{m}$}
\State {Compute ML loss $\mathcal{L}(\boldsymbol{\phi})$ according to eq.\ref{eqn:14}}
\State {Update neural network parameters $\boldsymbol{\phi}$ via backpropagation}
\Until {convergence to $\boldsymbol{\phi}^{*}$}
\State \emph{Inference (given observed or test data $\boldsymbol{x}_{obs}$):}
\State {Summarize the observed data by computing $\tilde{\boldsymbol{x}}_{obs} = h(\boldsymbol{x}_{obs},\boldsymbol{\phi}_{sum}^{*})$}
\For{$l = 1,...,L$}
\State {Sample $\boldsymbol{z}^{(l)} \sim \mathcal{N}_{d}(\boldsymbol{0},\boldsymbol{I})$}
\State {Compute $\boldsymbol{\theta}^{(l)} = f^{-1}(\boldsymbol{z}^{(l)};\tilde{\boldsymbol{x}}_{obs},\boldsymbol{\phi}_{inv})$}
\EndFor
\State {Use $\{\boldsymbol{\theta}^{(l)}\}_{l=1}^{L}$ to approximate the posterior $p(\boldsymbol{\theta}|\boldsymbol{x}_{obs})$}
\end{algorithmic}
\end{algorithm}

@@

\subsection*{Toy Example}

\subsection*{Example 1}

\subsection*{Example 2}

\subsection*{Example 3}

\subsection*{Example 4}

PNAS generally uses a two-column format averaging 67 characters, including spaces, per line. The maximum length of a Direct Submission research article is six pages and a Direct Submission Plus research article is ten pages including all text, spaces, and the number of characters displaced by figures, tables, and equations.  When submitting tables, figures, and/or equations in addition to text, keep the text for your manuscript under 39,000 characters (including spaces) for Direct Submissions and 72,000 characters (including spaces) for Direct Submission Plus.

\subsection*{Language-Editing Services}
Prior to submission, authors who believe their manuscripts would benefit from professional editing are encouraged to use a language-editing service (see list at www.pnas.org/site/authors/language-editing.xhtml). PNAS does not take responsibility for or endorse these services, and their use has no bearing on acceptance of a manuscript for publication.

>>>>

\dropcap{M}athematical models are formal descriptions of scientific theories allowing a clear and unambiguous way to formulate and test scientific hypotheses about probabilistic phenomena in a probabilistic world. In its most abstract form, a mathematical model is specified by a set of parameters $\boldsymbol{\theta}$ and a forward model $q$ mimicking the process by which manifest data $\boldsymbol{x}$ arise from latent parameters: 
\begin{align*}
\boldsymbol{x}= q(\boldsymbol{\theta}) \numberthis \label{eqn:1} 
\end{align*}
While $q$ can represent an arbitrarily complex process by an arbitrarily complicated expression, its functional form is usually guided by a well-founded theoretical framework. For instance, $q$ can be a stochastic differential equation describing the dynamics of single neurons in the brain, or a step-by-step biological algorithm dictating the rate of gene expression in certain cells. Thus, it is only through theoretical embedding that a meaningful interpretation in terms of some mechanism can be attached to the parameters of a mathematical model. Examples of mathematical models can be found in various scientific domains, e.g., genetics \cite{zappia2017splatter, beaumont2002approximate}, cognitive science \cite{palestro2018likelihood, usher2001time}, neuroscience \cite{hwang2018conditional, lueckmann2017flexible}, population dynamics \cite{wood2010statistical}, epidemiology \cite{hethcote2000mathematics}, just to name a few.

Once a mathematical model has been formulated, the next step consists of fitting the model to experimental or observational data and recovering the parameters of interest. However, estimating the parameters of a mathematical model from data can quickly become one of the most tenacious challenges in applications to real-world problems. It is also one of the most important ones to be tackled, since without reliable parameter estimation methods, it is impossible to test the utility of a model, regardless of its sophistication or theoretical appeal. Idealized parameter estimation involves computing the inverse (backward) model $\boldsymbol{\theta} = q^{-1}(\boldsymbol{x})$ exactly. However, due to noise, inherent stochasticity or loss of information, the inverse usually does not exist, so researchers need to resort to the sophisticated frameworks of Bayesian or frequentist inference.

@@

Likelihood-free methods aim at bypassing the above problems by resorting to a simulation-based approach to parameter estimation and model selection \cite{palestro2018likelihood, turner2014generalized}. A subset of likelihood-free methods includes approximate Bayesian computation (ABC) methods, which aim at preserving the advantages of Bayesian data analysis even when the likelihood function is intractable or practically impossible to compute \cite{turner2014generalized, sunnaaker2013approximate, csillery2010approximate}. ABC methods approximate the likelihood function by repeatedly sampling parameters from a pre-specified prior distribution $p(\boldsymbol{\theta})$ and then simulating multiple datasets by running the generative model $q(\boldsymbol{\theta})$ using the sampled parameters. Thus, the core ingredients of ABC methods are a prior on $\boldsymbol{\theta}$, and a generative model $q(\boldsymbol{\theta})$, usually specified as a function code in a general-purpose programming language \cite{csillery2010approximate, mertens2018abrox}.

@@

Recently, ideas from machine learning and deep learning research have entered the field of likelihood-free inference in an attempt to overcome some of the shortcomings of traditional methods \cite{radev2019towards, hwang2018conditional, mestdagh2018prepaid, raynal2018abc, jiang2017learning, lueckmann2017flexible, papamakarios2016fast}. The most common approach has been to cast the problem of parameter estimation as a supervised learning task. In this setting, a large dataset of the form $\boldsymbol{D} = \{h(\boldsymbol{x}^{(i)}), \boldsymbol{\theta}^{(i)}\}_{i=1}^{n}$ is created by repeatedly sampling from $p(\boldsymbol{\theta})$ and simulating an artificial dataset $\boldsymbol{x}$ by running $q(\boldsymbol{\theta})$ with the sampled parameters. Usually, the dimensionality of the simulated data is reduced by computing summary statistics  with a fixed summary function $h(\boldsymbol{x})$. Then, a supervised learning algorithm $f(h(\boldsymbol{x});\boldsymbol{\phi}) = \widehat{\boldsymbol{\theta}}$ with learnable parameters $\boldsymbol{\phi}$ (e.g., linear regression, random forest, neural network) is trained on the simulated data to later output an estimate of the true data generating parameters. Thus, $f(h(\boldsymbol{x});\boldsymbol{\phi})$ essentially attempts to “learn” the intractable inverse model $\boldsymbol{\theta} = q^{-1}(\boldsymbol{x})$.

@@

In the context of mathematical modeling and Bayesian inference, the target distribution is the posterior distribution of model parameters $p(\boldsymbol{\theta}|\boldsymbol{x})$ capturing our uncertainty about the numerical values of parameters given empirical data. We can leverage the fact that most mathematical models are generative in nature and as such can be used to perform multiple simulations of the process of interest. By specifying a prior distribution over the model parameters $p(\boldsymbol{\theta})$, one can generate arbitrarily large datasets of the form $\boldsymbol{D} = \{\boldsymbol{x}^{(i)}, \boldsymbol{\theta}^{(i)}\}_{i=1}^{n}$ and use a deep generative model to learn a probabilistic mapping from data to parameters. Thus, at inference time, one can condition the model on observed data $\boldsymbol{x}_{obs}$ and obtain samples $\boldsymbol{\theta}_{1}, \boldsymbol{\theta}_{2},...,\boldsymbol{\theta}_{L}$  approximating the posterior $p(\boldsymbol{\theta}|\boldsymbol{x}=\boldsymbol{x}_{obs})$ in the manner described above.

@@

The basic building block of a cINN is the affine coupling block (ACB) \cite{ardizzone2018analyzing, kingma2018glow, dinh2016density}. Each ACB consists of four separate fully connected neural networks denoted as $s_{1}(\cdot), s_{2}(\cdot), t_{1}(\cdot), t_{2}(\cdot)$. An ACB is specifically designed to be invertible, which means that in addition to a parametric mapping $f_{\boldsymbol{\phi}_{inv}}: \mathbb{R}^{d} \rightarrow \mathbb{R}^{d}$ it also learns the inverse mapping $f_{\boldsymbol{\phi}_{inv}}^{-1}: \mathbb{R}^{d} \rightarrow \mathbb{R}^{d}$ "for free". Denoting the input vector of $f_{\boldsymbol{\phi}_{inv}}$ as $\boldsymbol{u}$ and the output vector as $\boldsymbol{v}$, it follows that $f(\boldsymbol{u}; \boldsymbol{\phi}_{inv}) = \boldsymbol{v}$ and $f^{-1}(\boldsymbol{v}; \boldsymbol{\phi}_{inv}) = \boldsymbol{u}$. Invertibility is achieved by splitting the input vector into two parts $\boldsymbol{u} = (\boldsymbol{u}_{1}, \boldsymbol{u}_{2})$ and performing the following operations on the split input:
\begin{align*} 
\boldsymbol{v}_{1} &= \boldsymbol{u}_{1} \odot \exp(s_{1}(\boldsymbol{u}_{2})) + t_{1}(\boldsymbol{u}_{2}) \numberthis \label{eqn:2}  \\ 
\boldsymbol{v}_{2} &= \boldsymbol{u}_{2} \odot \exp(s_{1}(\boldsymbol{v}_{1})) + t_{1}(\boldsymbol{v}_{1}) \numberthis \label{eqn:3} 
\end{align*}
The outputs $\boldsymbol{v} = (\boldsymbol{v}_{1}, \boldsymbol{v}_{2})$ are then concatenated again and passed to the next ACB. The inverse operation is given by:
\begin{align*} 
\boldsymbol{u}_{2} &= (\boldsymbol{v}_{2} - t_{2}(\boldsymbol{v}_{1})) \odot \exp(-s_{2}(\boldsymbol{v}_{1})) \numberthis \label{eqn:4}  \\ 
\boldsymbol{u}_{1} &= (\boldsymbol{v}_{1} - t_{1}(\boldsymbol{u}_{2})) \odot \exp(-s_{1}(\boldsymbol{u}_{2})) \numberthis \label{eqn:5} 
\end{align*}
An additional property of this design, which becomes relevant later for optimization, is that the operations of the ACB have tractable, and cheaply computable Jacobians (strictly upper or lower triangular matrices). Furthermore, the internal networks $s_{1}(\cdot), s_{2}(\cdot), t_{1}(\cdot), t_{2}(\cdot)$ can be represented by arbitrarily complex neural networks, which themselves need not be invertible, since they are only ever evaluated in the forward direction during both the forward and the inverse pass through the ACB. To ensure that the model is powerful enough to represent complicated distributions, we chain multiple ACBs, so that the output of each ACB becomes the input of the next. In this way, the whole chain remains invertible from the first input to the last output and can be viewed as a single function parameterized by trainable parameters $\boldsymbol{\phi}_{inv}$.

@@

\subsection*{Training the Networks}
We train all cINNs and summary networks described in this paper jointly via backpropagation. For all following experiments, we use the Adam optimizer  with a starter learning rate of $10^-3$ and an exponential decay rate of $.95$. We set the weight regularization parameter $\tau$ to a value of $10^{-5}$. We perform $100000$ iterations (network updates) for each of the examples in this paper, and report the results on the trained network. Note, that we did not perform an extensive search for optimal values of the cINN hyperparameter but use a cINN with 10 ACBs for each example in this paper (see \textbf{SI} for details of the cINN).
Concerning the data generation step, we use two training approaches. The first follows the classical approximate Bayesian computation approach to create a large “reference table” or grid of the form $\boldsymbol{D} = \{\boldsymbol{x}^{(i)}, \boldsymbol{\theta}^{(i)}\}_{i=1}^{n}$. The reference table is then used as training data for the neural network and training continues for a pre-specified number of epochs through the entire reference table. A separate validation dataset is eventually used to assess the performance of the network. This training approach separates the simulation from the training phase but can incur large memory overhead, as the reference table must be stored on disk and then loaded in chunks or in its entirety into memory. The second approach follows a different strategy, which is used mainly in the field of active learning. Correspondingly, a dataset, or a batch of datasets, is created on the fly and then passed through the neural network. This training regime has the advantage that the network never “experiences” the same input data twice. Moreover, training can continue as long as the network keeps improving (i.e., the loss keeps decreasing), since overfitting in the classical sense is nearly impossible. However, if the simulations are computationally expensive and researchers need to experiment with multiple models, it might be beneficial to switch to the first regime, since simulation and training in the active learning regime are tightly intertwined. TODO - Describe for which!

@@

On an abstract level, our method requires three key ingredients: 1) a mathematical process model $q(\boldsymbol{\theta})$ capable of simulating data $\boldsymbol{x}$; 2) a prior distribution over the model parameters $p(\boldsymbol{\theta})$ encoding our prior beliefs about plausible parameter values; 3) an invertible neural network $f_{\boldsymbol{\theta}}$ capable of approximating a large enough family of probability distributions (see Figure 2). In practice, a chain of up to 10 ACBs should suffice to learn most distributions encountered in the life sciences, since they tend to be unimodal and relatively simple (in contrast to the distribution of natural images or words in a spoken language). From these three ingredients, a universal and reusable sampler can be designed for likelihood-free Bayesian estimation of both tractable and intractable mathematical models. \textbf{Algorithm} \ref{Algorithm:algo} describes the essential steps of the method using an arbitrary summary network and employing the active learning training regime.

\begin{algorithm}
\caption{Bayesian likelihood-free inference with invertible neural networks}\label{Algorithm:algo}
\begin{algorithmic}[1]
\State \emph{Training (via active learning):}
\Repeat
\State {Sample a batch of $\{\boldsymbol{\theta}^{(i)}\}_{i=1}^{m}$ from prior $p(\boldsymbol{\theta})$}
\State {Simulate a batch of datasets $\{\boldsymbol{x}^{(i)}\}_{i=1}^{m}$ by running $\boldsymbol{x}^{(i)} = q(\boldsymbol{\theta}^{(i)})$ for $i=1,...,m$}
\State {Pass $\{\boldsymbol{x}^{(i)}\}_{i=1}^{m}$ through summary network $h(\boldsymbol{x}^{(i)};\boldsymbol{\phi}_{sum})$ to obtain $\{\tilde{\boldsymbol{x}}^{(i)}\}_{i=1}^{m}$}
\State {Pass $\{\boldsymbol{\theta}^{(i)}\}_{i=1}^{m}$ and $\{\tilde{\boldsymbol{x}}^{(i)}\}_{i=1}^{m}$ through cINN $f(\boldsymbol{\theta}^{(i)};\boldsymbol{x}^{(i)},\boldsymbol{\phi}_{inv})$ to obtain $\{\boldsymbol{z}^{(i)}\}_{i=1}^{m}$}
\State {Compute ML loss $\mathcal{L}(\boldsymbol{\phi})$ according to eq.\ref{eqn:14}}
\State {Update neural network parameters $\boldsymbol{\phi}$ via backpropagation}
\Until {convergence to $\boldsymbol{\phi}^{*}$}
\State \emph{Inference (given observed or test data $\boldsymbol{x}_{obs}$):}
\State {Summarize the observed data by computing $\tilde{\boldsymbol{x}}_{obs} = h(\boldsymbol{x}_{obs},\boldsymbol{\phi}_{sum}^{*})$}
\For{$l = 1,...,L$}
\State {Sample $\boldsymbol{z}^{(l)} \sim \mathcal{N}_{d}(\boldsymbol{0},\boldsymbol{I})$}
\State {Compute $\boldsymbol{\theta}^{(l)} = f^{-1}(\boldsymbol{z}^{(l)};\tilde{\boldsymbol{x}}_{obs},\boldsymbol{\phi}_{inv})$}
\EndFor
\State {\textbf{end for}}
\State {Use $\{\boldsymbol{\theta}^{(l)}\}_{l=1}^{L}$ to approximate the posterior $p(\boldsymbol{\theta}|\boldsymbol{x}_{obs})$}
\end{algorithmic}
\end{algorithm}

@@

\subsection*{Toy Example – Bayesian Regression}

As a proof-of-concept, we demonstrate the utility of our method in recovering the true analytic posteriors of the regression coefficients of a conjugate Bayesian regression model. To set the setting, assume we have observed a dataset $\boldsymbol{D} = \{(\boldsymbol{x}^{(i)},y^{(i)})\}_{i=1}^{n}$ with $\boldsymbol{x} \in \mathbb{R}^{d}$ and $y \in \mathbb{R}$ We model each $y^{(i)}$ as being conditionally Gaussian given $x^{(i)}$, i.e.,: $y^{(i)} \sim \mathcal{N}(\boldsymbol{\theta}^{T}\boldsymbol{x}^{(i)},a^{-1})$ where $\boldsymbol{\theta} \in \mathbb{R}^{d}$ and $a$ is the precision (inverse noise variance, $a \equiv 1/\sigma_{y}^{2}$). We place a $d$-dimensional diagonal Gaussian prior on the regression coefficients centered at $0$: $\boldsymbol{\theta} \sim \mathcal{N}_{d}(\boldsymbol{0},b^{-1}\boldsymbol{I})$ where $b$ is the precision of the prior ($b \equiv 1/\sigma_{\boldsymbol{\theta}}^{2}$). Thus, the likelihood $p(\boldsymbol{D}|\boldsymbol{\theta})$ admits the following proportionality:
\begin{align*}
p(\boldsymbol{D}|\boldsymbol{\theta}) \propto \exp\left(-\frac{a}{2}\left(y-\boldsymbol{X}\boldsymbol{\theta}\right)^{T}\left(y - \boldsymbol{X}\boldsymbol{\theta}\right)\right) \numberthis 
\end{align*}
where $\boldsymbol{X}$ denotes the design matrix (with all $\boldsymbol{x}^{(i)}$ stacked row-wise). Since the prior of $\boldsymbol{\theta}$ is conjugate to the likelihood (both are Gaussian distributions), the posterior of $\boldsymbol{\theta}$ is also Gaussian, given by:
\begin{align*}
p(\boldsymbol{\theta}|\boldsymbol{D}) \propto \exp\left(-\frac{a}{2}\left(y-\boldsymbol{X}\boldsymbol{\theta}\right)^{T}\left(y - \boldsymbol{X}\boldsymbol{\theta}\right)-\frac{b}{2}\boldsymbol{\theta}^{T}\boldsymbol{\theta}\right) \numberthis 
\end{align*}
Therefore, the posterior has the form $p(\boldsymbol{\theta}|\boldsymbol{D}) = \mathcal{N}_{d}(\boldsymbol{\theta}|\boldsymbol{\mu},\boldsymbol{\Lambda}^{-1})$ where $\boldsymbol{\Lambda}$ denotes the posterior precision matrix (inverse covariance matrix), $\boldsymbol{\mu}$ the posterior mean vector, which are computed as follows:
\begin{align*}
\boldsymbol{\Lambda} &= a\boldsymbol{X}^{T}\boldsymbol{X} + b\boldsymbol{I} \numberthis \\
\boldsymbol{\mu} &= a\boldsymbol{\Lambda}^{-1}\boldsymbol{X}^{T}\boldsymbol{y} \numberthis
\end{align*} 
Thus, for known $a$ and $b$, the posterior can be easily computed. Even though in real-world applications $a$ is usually not known and a hierarchical model is used instead, the current example is good for proving the utility of our method.

For the following application of our method, we set $d=4$, and $a=b=1$. The design matrices for each iteration contain a variable number $n$ of \textit{i.i.d.} data points drawn from a unit Gaussian $\mathcal{N}_{4}(\boldsymbol{0},\boldsymbol{I})$. The number of trials is drawn from a uniform distribution $n \sim \mathcal{U}(50, 500)$ at each training iteration (Lines $2$-$9$ of \textbf{Algorithm} \ref{Algorithm:algo}).

The results on the toy example are depicted in Figure.XXX. The approximate posterior means show negligible deviations from the analytic posterior means as quantified by very small $NRMSE$s (as small as XXX) and very high $R^{2}$ (as high as $1.0$) over all numbers of data points used for training. This suggests near-perfect reconstruction of the true posterior means.

\subsection*{Example 1 - The Ricker Model}
Discrete population dynamics models describe how the number of individuals in a population changes over discrete units of time \cite{wood2010statistical}. In particular, the Ricker model describes the number of individuals $x_{t+1}$ in generation $t+1$ as a function of the number of individuals in the previous generation $t$ by the following non-linear equation:
\begin{align*}
N_{t+1} &= rN_{t}\exp(-N_{t} + \epsilon_{t}) \numberthis \\
x_{t} &\sim Pois(\rho N_{t}) \numberthis
\end{align*}
for $t = 1,...,T$ where $N_{t}$ is the expected number of individuals at time $t$, $r$ is the growth rate, $\rho$ is a scaling parameter and $\epsilon_{t} \sim \mathcal{N}(0, \sigma^{2})$ is random Gaussian noise \cite{mestdagh2018prepaid}. The Ricker model has no known likelihood function and is thus a suitable candidate for likelihood-free inference. The parameter estimation task is thus to recover $\boldsymbol{\theta}=(\rho,r,\sigma)$ from the observed one-dimensional time-series data $\boldsymbol{x}=(x_{1},x_{2},...,x_{T})$ where each $x_{t} \in \mathbb{N}$.

During training of the networks, we simulate time-series from the Ricker model with varying lengths. The number of time points $T$ is drawn from a uniform distribution $T \sim \mathcal{U}(100, 500)$ at each training iteration (see the \textbf{SI} for more details about the simulation).

The results on the Ricker model are depicted in Figure XXX... TODO

\subsection*{Example 2 - The Lévy-Flight Model}
Evidence accumulator models (EAMs) describe (perceptual) decision making by a set of neurocognitively motivated parameters CITE. EAMs are most often applied to choice reaction times (RT) data to obtain an estimate of the neurocognitive processes governing observed RT distributions in human and animal participants. Most EAM flavors share four underlying assumptions: \textit{i)} information about a stimulus (response option) is accumulated continuously through time; \textit{ii)} stochasticity in the form of noisy accumulation ensures variability; \textit{iii)} empirical response times can be decomposed into a decision time component and a non-decision time component accounting for pre-decisional perceptual (encoding time) and post-decisional motor processes (response execution); and \textit{iv)} a decision is met when the activation of an accumulator exceeds a threshold. In its most general formulation, the forward model of EAMs takes the form of a stochastic differential equation given by:
\begin{align*} 
dx = vdt + cd\xi  \numberthis \label{eqn:stillno}
\end{align*}
where $dx$ denotes a change in activation of an accumulator, $v$ denotes the average speed of information processing (often termed the drift rate), and $d\xi$ represents a stochastic additive component, usually modeled as following a Gaussian distribution centered around $0$: $d\xi \sim \mathcal{N}(0, c^{2})$.

EAMs are particularly amenable for likelihood-free inference, since most members of the family turn out to be intractable. This intractability has precluded many interesting applications and empirically driven model refinements. Here, we apply our method to estimate the parameters of the recently proposed Lévy-Flight Model (LFM, cite). The LFM assumes an \textit{alpha-stable} noise distribution of the evidence accumulation process in order to model "jumps" in the decision process. The LFM does not have a known likelihood and a fully Bayesian treatment of the model is still missing from the literature. The forward equation of the LFM is given by:
\begin{align*}
dx &= vdt + \xi dt^{1/\alpha} \numberthis \label{eqn:stillno2} \\
\xi &\sim AlphaStable(\alpha,0,1,0) \numberthis \label{eqn:stillno3}
\end{align*}
The LFM has three additional parameters: a threshold $a$ determining the amount of evidence needed for the termination of a decision process; a relative starting point, $zr$, determining the amount of starting evidence available to the accumulator before the actual decision alternatives are presented; and an additive non-decision time $t_{0}$.

During training of the networks, we simulate RT data from two experimental conditions with two different drift rates (see \textbf{SI} for details of the simulation). The parameter estimation task is thus to recover the parameters $\boldsymbol{\theta} = (v_{0}, v_{1}, a, t_{0}, zr, \alpha)$ from two-dimensional \textit{i.i.d.} RT data $\boldsymbol{X} = (\boldsymbol{x}_{1},\boldsymbol{x}_{2},...,\boldsymbol{x}_{n})$ containing variable number or RT trials. The number of trials is drawn from a uniform distribution $n \sim \mathcal{U}(100, 1000)$ at each training iteration.
